Abstractions for Type-Level Programming

### 2022-03-19 slides.tex
Slides

%% Minted setup (syntax highlighting)
\usepackage[outputdir=latex.out]{minted}
\usemintedstyle{vs}
\setminted{fontsize=\footnotesize}
\newminted[code]{scala}{gobble=0, escapeinside=@@}
\newenvironment{slide}[2][]
  {\begin{frame}[fragile,environment=slide,#1]{#2}}
  {\end{frame}}
\newcommand\makemintedshortinline[2]{%
  \catcode`#2=13
  \begingroup%
  \lccode`\~=`#2
  \lowercase{\endgroup\protected\def~{\mintinline{#1}~}}%
}
\renewcommand{\fcolorbox}[4][]{#4}

@@

\usepackage{xspace}
\usepackage{mathtools}
\def\C{\normalfont\text{C}}
\def\<:{\allowbreak\operatorname{\normalfont\text{<}:}\allowbreak}
\def\Fsub{\normalfont{F}\ensuremath{_{\<:}}\xspace}

@@

\usepackage{xcolor}
\definecolor{typeColor}{HTML}{2B91B0}
\definecolor{implColor}{HTML}{57B5E8}
\definecolor{singColor}{HTML}{009E73}
\definecolor{mtpeColor}{HTML}{9400D4}
\newcommand{\tp}{\color{typeColor}}

@@

\onslide<3>
\begin{code}
dependent
def int2str(x: Int)
  : { if (x @\tp{}== 0@) "zero" @\tp{}else@ "one" }
  =   if (x == 0) "zero" else "one"
\end{code}

\onslide<4-5>
\begin{code}
dependent
def int2str(x: Int)
  : If[x.type @\tp{}== 0@, "zero", "one"]
  =   if (x == 0) "zero" else "one"
\end{code}

\onslide<6>
\begin{code}
dependent
def int2str(x: Int)
  : If[x.type @\tp{}== 0@, "zero", String]
  =   if (x == 0) "zero" else readString()
\end{code}

\onslide<7>
\begin{code}
dependent
def int2str(x: Int)
  : { if (x @\tp{}== 0@) "zero" @\tp{}else@ (_: String) }
  =   if (x == 0) "zero" else readString()
\end{code}
\end{overprint}
\medskip
\pause\pause\pause\pause

@@

\begin{slide}{List concatenation with generalized singleton types}
\begin{code}
dependent def concat(xs: List, ys: List) <: List =
  xs match
    case x :: xs => x :: concat(xs, ys)
    case Nil => ys
\end{code}
\pause
\begin{overprint}
\onslide<1-2>
\begin{code}
dependent val l1 = "A" :: Nil
dependent val l2 = "B" :: Nil
dependent val l3 = concat(l1, l2)
l3: { "A" :: "B" :: Nil }
\end{code}

@@

We evaluate type tests using subtyping and type disjointness:
\begin{code}
eval(@\tp{}x.asInstanceOf@[Foo]) = true   iff  x.type <: Foo
eval(@\tp{}x.asInstanceOf@[Foo]) = false  iff  disj(x.type, Foo)
\end{code}
\end{slide}

\begin{slide}{Type evaluation: pattern matching example}
\begin{overprint}
\onslide<1>
\begin{code}
dependent val foo(x: Any) =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\onslide<2>
\begin{code}
dependent val foo(x: Any): {
  x match
    case s: String => @\tp{}s@
    @\tp{}case i: Int@ => @\tp{}i+1@
} =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\onslide<3->
\begin{code}
dependent val foo(x: Any): {
  if (x.isInstanceOf[String]) x.asInstanceOf[String]
  @\tp{}else if @(@\tp{}x.isInstanceOf@[@\tp{}Int@]) @\tp{}x.asInstanceOf@[Int] @\tp{}+ 1@
  @\tp{}else throw new@ MatchError()
} =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\end{overprint}
\bigskip
\begin{overprint}
\onslide<4>
|foo(readString())|
\onslide<5>
|foo(readString()): { foo(_: String) }|
\onslide<6>
|foo(readString()): { (_: String).asInstanceOf[String] }|
\onslide<7>
|foo(readString()): { (_: String) }|
\onslide<8>
|foo(readString()): String|
\end{overprint}
\end{slide}

\begin{slide}{Generalized Singletons: Conclusion}
TODO
\end{slide}

@@

FM(C, $\Psi$, $\Xi$)
\begin{itemize}
  \item $\C \hspace{2.3pt} \coloneqq$ set of classes
  \item $\Psi \coloneqq$ class inheritance relation
  \item $\Xi \hspace{1.1pt} \coloneqq$ class disjointness relation
\end{itemize}

For example:

|class A|                    -> C={A}, $\Psi$={}, $\Xi$={}
|class A; class B extends A| -> C={A, B}, $\Psi$={(A, B)}, $\Xi$={}
|class A; class D|           -> C={A, D}, $\Psi$={}, $\Xi$={(A, D)}
|class A; trait T|           -> C={A, T}, $\Psi$={}, $\Xi$={}
\end{slide}

@@

\begin{slide}{System FM's disjointness relation}
\includegraphics[width=\textwidth]{figures/FMDisjointness.pdf}
\end{slide}

\begin{slide}{System FM's new subtyping rules}
\includegraphics[width=\textwidth]{figures/FMSubtyping.pdf}
\end{slide}

\begin{slide}{System FM's type safety}
theorems

two versions of the proof, pen and paper and Coq
\end{slide}

\begin{slide}{Structure of the type safety proof}
\includegraphics[width=\textwidth]{figures/FMProofStructure.pdf}
\end{slide}

\begin{slide}{Limitations}
- type binders
- higher-kinded types
- recursion
\end{slide}

\begin{slide}{Empty Types}
\begin{code}
type M[X] = X match
  case Int => String
  case String => Int
class C:
  type X
  def f(bad: M[X & String]): Int = bad
class D extends C:
  type X = Int
\end{code}
\end{slide}

\begin{slide}{Variance}
\begin{code}
F[+T]: T1 <: T2 => F[T1] <: F[T2]
G[-T]: T1 <: T2 => G[T1] >: G[T2]

case class Tuple2[+T1, +T2](_1: T1, _2: T2)
\end{code}
\end{slide}

@@

\begin{slide}{Type-Safe Regular Expressions}
\timeline{80}{80}{80}{0}{80}
\end{slide}

\begin{slide}{Type-Safe Regular Expressions (2022)}
\begin{code}
val rational = new Regex("(\\d+)\\.?(\\d+)?")
rational.unapply("3.1415"): Option[(String, Option[String])]
\end{code}
\pause
\begin{code}
type GroupsOf[P <: String] = ...

\end{code}
\end{slide}

>>>>

%% Minted setup (syntax highlighting)
\usepackage[outputdir=latex.out]{minted}
\usemintedstyle{vs}
\setminted{fontsize=\footnotesize}
\newminted[code]{scala}{gobble=0, escapeinside=@@}
\newenvironment{slide}[2][]
  {\begin{frame}[fragile,environment=slide,#1]{#2}}
  {\end{frame}}
\newcommand\makemintedshortinline[2]{%
  \catcode`#2=13
  \begingroup%
  \lccode`\~=`#2
  \lowercase{\endgroup\protected\def~{\mintinline[escapeinside=@@]{#1}~}}%
}
\renewcommand{\fcolorbox}[4][]{#4}

@@

\usepackage{xspace}
\usepackage{mathtools}
\def\A{\normalfont\text{A}}
\def\B{\normalfont\text{B}}
\def\C{\normalfont\text{C}}
\def\D{\normalfont\text{D}}
\def\T{\normalfont\text{T}}
\def\<:{\allowbreak\operatorname{\normalfont\text{<}:}\allowbreak}
\def\Fsub{\normalfont{F}\ensuremath{_{\<:}}\xspace}

@@

\usepackage{xcolor}
\definecolor{typeColor}{HTML}{2B91B0}
\definecolor{implColor}{HTML}{57B5E8}
\definecolor{singColor}{HTML}{009E73}
\definecolor{mtpeColor}{HTML}{9400D4}
\newcommand{\tp}[1]{\color{typeColor}#1}

@@

\onslide<3>
\begin{code}
dependent
def int2str(x: Int)
  : { if (x @\tp{== 0}@) "zero" @\tp{else}@ "one" }
  =   if (x == 0) "zero" else "one"
\end{code}

\onslide<4-5>
\begin{code}
dependent
def int2str(x: Int)
  : If[x.type @\tp{== 0}@, "zero", "one"]
  =   if (x == 0) "zero" else "one"
\end{code}

\onslide<6>
\begin{code}
dependent
def int2str(x: Int)
  : If[x.type @\tp{== 0}@, "zero", String]
  =   if (x == 0) "zero" else readString()
\end{code}

\onslide<7>
\begin{code}
dependent
def int2str(x: Int)
  : { if (x @\tp{== 0}@) "zero" @\tp{else}@ (_: String) }
  =   if (x == 0) "zero" else readString()
\end{code}
\end{overprint}
\medskip
\pause\pause\pause\pause

@@

\begin{slide}{"Precise mode" of type inference: list concatenation example}
\begin{code}
dependent def concat(xs: List, ys: List) <: List =
  xs match
    case x :: xs => x :: concat(xs, ys)
    case Nil => ys
\end{code}
\pause
\begin{overprint}
\onslide<1-2>
\begin{code}
dependent val l1 = "A" :: Nil
dependent val l2 = "B" :: Nil
dependent val l3 = concat(l1, l2)
l3: { "A" :: "B" :: Nil }
\end{code}

@@

We evaluate type tests using subtyping and type disjointness:
\begin{code}
eval(@\tp{x.asInstanceOf}@[Foo]) = true   iff  x.type <: Foo
eval(@\tp{x.asInstanceOf}@[Foo]) = false  iff  disj(x.type, Foo)
\end{code}
\end{slide}

\begin{slide}{Type evaluation: pattern matching example}
\begin{overprint}
\onslide<1>
\begin{code}
dependent val foo(x: Any) =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\onslide<2>
\begin{code}
dependent val foo(x: Any): {
  x match
    case s: String => @\tp{s}@
    @\tp{case i: Int}@ => @\tp{i+1}@
} =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\onslide<3->
\begin{code}
dependent val foo(x: Any): {
  if (x.isInstanceOf[String]) x.asInstanceOf[String]
  @\tp{else if }@(@\tp{x.isInstanceOf}@[@\tp{Int}@]) @\tp{x.asInstanceOf}@[Int] @\tp{+ 1}@
  @\tp{else throw new}@ MatchError()
} =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\end{overprint}
\bigskip

\begin{overprint}
\onslide<4>
|foo(42): { foo(42) }|
\onslide<5>
|foo(42): { 42.asInstanceOf[Int] + 1 }|
\onslide<6>
|foo(42): { 42 + 1 }|
\onslide<7>
|foo(42): { 43 }|
\onslide<8->
|foo(42): 43|
\end{overprint}

\begin{overprint}
\onslide<9>
|foo(readString())|
\onslide<10>
|foo(readString()): { foo(_: String) }|
\onslide<11>
|foo(readString()): { (_: String).asInstanceOf[String] }|
\onslide<12>
|foo(readString()): { (_: String) }|
\onslide<13>
|foo(readString()): String|
\end{overprint}
\end{slide}

\begin{slide}{Generalized Singletons: Recap}

We proposed the following 3 changes:
\begin{enumerate}
  \item lift a subset of Scala's language constructs to the type level
  \item add a "precise mode" of type inference
  \item evaluate those types during subtyping
\end{enumerate}
\end{slide}

TODO
\end{slide}

@@

FM(C, $\Psi$, $\Xi$)
\begin{itemize}
  \item $\C \hspace{2.3pt} \coloneqq$ set of classes
  \item $\Psi \coloneqq$ class inheritance relation
  \item $\Xi \hspace{1.1pt} \coloneqq$ class disjointness relation
\end{itemize}
\pause
For example:

\begin{tabular}{@{}llll@{}}
|class A|          &$\C = \{\A\}$     & $\Psi = \varnothing$  & $\Xi = \varnothing$\\
|class A; class B extends @\tp{A}@| &$\C = \{\A, \B\}$ & $\Psi = \{(\B, \A)\}$ & $\Xi = \varnothing$\\
|class A; class D| &$\C = \{\A, \D\}$ & $\Psi = \varnothing$  & $\Xi = \{(\A, \D)\}$\\
|class A; trait T| &$\C = \{\A, \T\}$ & $\Psi = \varnothing$  & $\Xi = \varnothing$\\
\end{tabular}
\end{slide}

@@

\begin{slide}{System FM's pattern matching evaluation}
Given a concrete scrutinee, we evaluate match expressions with a few lookups in the inheritance relation:
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/FMEMatch.pdf}
\end{center}
\end{slide}

\begin{slide}{System FM's disjointness relation}
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/FMDisjointness.pdf}
\end{center}
\end{slide}

\begin{slide}{System FM's match types evaluation rule (subtyping)}
\includegraphics[width=\textwidth]{figures/FMSMatch.pdf}
\end{slide}

\begin{slide}{System FM's type safety}
We show type safety through the progress and preservation.

Our proof comes in two versions:
\begin{enumerate}
  \item Pen and paper (\textasciitilde30 pages)
  \item Coq mechanization (\textasciitilde6000 LOC)
\end{enumerate}
\end{slide}

\begin{slide}{Structure of the type safety proof}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/FMProofStructure.pdf}
\end{center}
\end{slide}

\begin{slide}{Empty Types}
Our system does not like empty types.

|Nothing|, in particular, is both subtype and disjoint from every type.

Intersections require special care.

\pause

\begin{code}
type M[X] = X match
  case Int => String
  case String => Int

class C:
  type X
  def f(bad: M[X & String]): Int = bad

class D extends C:
  type X = Int
\end{code}
\end{slide}

\begin{slide}{Variance}
At first glance, we can't prove disjointness when variance is involved.

|F[+T]:| \quad |T1 <: T2| \quad implies \quad |F[T1] <: F[T2]|\\
|G[-T]:| \quad |T1 <: T2| \quad implies \quad |G[T1] >: G[T2]|
\pause

We make an exception for covariant type that are used as class fields:
\begin{code}
case class Some[+A](value: A)
\end{code}

|Some[String]| and |Some[Int]| are disjoint, since there is no runtime value of type |Some[Nothing]|.

\end{slide}

@@

\begin{slide}{Thank you!}
\timeline{0}{0}{0}{0}{0}
\end{slide}

\appendix

\begin{slide}{Type-Safe Regular Expressions (simplified)}
\begin{code}
val rational = new Regex("(\\d+)\\.(\\d+)")
rational.unapply("3.1415"): Option[(String, String)]

type GroupsOf[P <: String] =
  Reverse[Loop[P, 0, Length[P], EmptyTuple]]

type Loop[P, Lo, Hi, Acc <: Tuple] =
  Lo match
    case Hi => Acc
    case _ => CharAt[P, Lo] match
      case '('  => Loop[P, Lo + 1, Hi, String *: Acc]
      case '\\' => Loop[P, Lo + 2, Hi, Acc]
      case _ => Loop[P, Lo + 1, Hi, Acc]
\end{code}
\end{slide}
